---
title: "Movies project markdown"
author: "Hassan Ali"
date: "2025-06-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
library(scales)
```

#  🎬 Data Preparation

## Read and Inspect the Data

```{r }
movie_data <- read_csv("Movie Data Starter Project - Movie Data.csv")
head(movie_data)
sort(colSums(is.na(movie_data)), decreasing = TRUE)

```

## Clean the Data

```{r}
clean_movie_data <- movie_data %>% 
  select(where(~ sum(is.na(.)) <= 20))
head(clean_movie_data)
```

# 🎭 Genre-wise Analysis
## Number of Movies per Genre

```{r Number of Movies per Genre}
No_of_movies_per_genry <- clean_movie_data %>% 
  group_by(Genre_1) %>% 
  summarise(No_of_movies = n())

ggplot(No_of_movies_per_genry, aes(x = Genre_1, y = No_of_movies)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of Movies Per Genre", x = "Genre", y = "No of Movies") +
  theme_minimal()

```


## Total Box Office Revenue per Genre

```{r Total Box Office Revenue per Genre}
revenue_by_genre <- clean_movie_data %>%
  group_by(Genre_1) %>%
  summarise(`Box office revenue` = round(sum(`Box Office Revenue`) / 1e6, 1))

ggplot(revenue_by_genre, aes(x = Genre_1, y = `Box office revenue`)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(`Box office revenue`, "M")), size = 3.5, vjust = -0.5) +
  labs(title = "Box Office Revenue per Genre", x = "Genre", y = "Revenue (Millions)") +
  theme_minimal()

```

## Budget per Genre

```{r Budget per Genre}
budget_by_genre <- clean_movie_data %>%
  group_by(Genre_1) %>%
  summarise(Budget = round(sum(Budget) / 1e6, 1))

ggplot(budget_by_genre, aes(x = Genre_1, y = Budget)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(Budget, "M")), size = 3.5, vjust = -0.5) +
  labs(title = "Total Budget per Genre", x = "Genre", y = "Budget (Millions)") +
  coord_flip() +
  theme_minimal()

```

# 💸 Profit & Loss Analysis

## Top 10 Loss-Making Movies

```{r Top 10 Loss-Making Movies}
loss_movies <- clean_movie_data %>%
  mutate(Profit_loss = `Box Office Revenue` - Budget,
         Loss_Amount = -Profit_loss) %>%
  filter(Profit_loss < 0) %>%
  arrange(desc(Loss_Amount)) %>%
  slice_head(n = 10)

loss_movies %>%
  mutate(`Movie Title` = fct_reorder(`Movie Title`, Loss_Amount)) %>%
  ggplot(aes(x = `Movie Title`, y = Loss_Amount)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(Loss_Amount / 1e6, 1), "M")),
            size = 3.5, hjust = -0.1) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Top 10 Loss-Making Movies", x = "Movie", y = "Loss (Millions)") +
  coord_flip() +
  theme_minimal()

```

## Top 10 Profit-Making Movies

```{r Top 10 Profit-Making Movies}
profit_movies <- clean_movie_data %>%
  mutate(Profit_loss = `Box Office Revenue` - Budget)

top10_profit <- profit_movies %>%
  arrange(desc(Profit_loss)) %>%
  slice_head(n = 10) %>%
  mutate(`Movie Title` = fct_reorder(`Movie Title`, Profit_loss))

ggplot(top10_profit, aes(x = `Movie Title`, y = Profit_loss)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(Profit_loss / 1e6, 1), "M")),
            size = 3.5, hjust = -0.1) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Top 10 Profit-Making Movies", x = "Movie", y = "Profit (Millions)") +
  coord_flip() +
  theme_minimal()

```

# 💰 Budget Insights

## Average Cost per Genre
```{r Average Cost per Genre}
avg_cost_for_type <- clean_movie_data %>%
  group_by(Genre_1) %>%
  summarise(Avg_cost = round(mean(Budget), 2)) %>%
  mutate(Genre_1 = fct_reorder(Genre_1, Avg_cost))

ggplot(avg_cost_for_type, aes(x = Genre_1, y = Avg_cost)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(Avg_cost / 1e6, 1), "M")), size = 3.5, hjust = -0.1) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Average Movie Budget per Genre", x = "Genre", y = "Average Budget (Millions)") +
  coord_flip() +
  theme_minimal()
```

## Most Expensive Movie by Genre

```{r Most Expensive Movie by Genre}
most_expensive_movies <- clean_movie_data %>%
  group_by(Genre_1) %>%
  slice_max(order_by = Budget, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(Budget))

most_expensive_movies %>%
  mutate(Genre_1 = fct_reorder(Genre_1, Budget)) %>%
  ggplot(aes(x = Genre_1, y = Budget)) +
  geom_bar(stat = "identity", fill = "tomato") +
  geom_text(aes(label = paste0(`Movie Title`, " (", round(Budget / 1e6, 1), "M)")),
            hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = label_comma()) +
  labs(title = "Most Expensive Movie by Genre", x = "Genre", y = "Budget (Millions)") +
  coord_flip() +
  theme_minimal()


```

# 🎬 Director Performance Analysis

## Top 10 Directors with Most Flops

```{r Top 10 Directors with Most Flops}
director_outcomes <- clean_movie_data %>%
  mutate(Profit_loss = `Box Office Revenue` - Budget,
         Outcome = if_else(Profit_loss > 0, "Successful", "Flop")) %>%
  group_by(Director_1, Outcome) %>%
  summarise(movie_count = n(), .groups = "drop")

total_counts <- director_outcomes %>%
  group_by(Director_1) %>%
  summarise(total_movies = sum(movie_count), .groups = "drop")

flop_counts <- director_outcomes %>%
  filter(Outcome == "Flop") %>%
  arrange(desc(movie_count)) %>%
  slice_head(n = 10) %>%
  select(Director_1)

top_directors <- director_outcomes %>%
  filter(Director_1 %in% flop_counts$Director_1) %>%
  left_join(total_counts, by = "Director_1")

ggplot(top_directors, aes(x = reorder(Director_1, -total_movies), y = movie_count, fill = Outcome)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Directors with Most Flops", x = "Director", y = "Number of Movies", fill = "Outcome") +
  coord_flip() +
  theme_minimal()

```

## Top 10 Directors with Most Successful Movies

```{r Top 10 Directors with Most Successful Movies}
top_successful_directors <- director_outcomes %>%
  filter(Outcome == "Successful") %>%
  arrange(desc(movie_count)) %>%
  slice_head(n = 13) %>%
  select(Director_1, successful_count = movie_count)

top_directors <- director_outcomes %>%
  filter(Director_1 %in% top_successful_directors$Director_1) %>%
  left_join(top_successful_directors, by = "Director_1")

ggplot(top_directors, aes(x = reorder(Director_1, -successful_count), y = movie_count, fill = Outcome)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Directors with Most Successful Movies", x = "Director", y = "Number of Movies", fill = "Outcome") +
  coord_flip() +
  theme_minimal()

```

# 📅 Trend Over Time

## Revenue & Budget by Year:

Analyze how total budget or revenue trends have changed over the years.

```{r Total Budget over time}
# Parse date-time and extract year
yearly_data <- clean_movie_data %>%
mutate(Release_Date_Parsed = mdy_hm(`Release Date`)) %>%
mutate(Year = year(Release_Date_Parsed)) %>% 
  group_by(Year) %>%
  summarise(
    Total_Budget = round(sum(Budget, na.rm = TRUE) / 1e6, 1),
    Total_Revenue = round(sum(`Box Office Revenue`,na.rm = TRUE)/1e6,1),
    Movie_count= n()
  ) %>%
  arrange(Year)

# Plot 1: Total Budget
ggplot(yearly_data, aes(x = Year, y = Total_Budget)) +
  geom_line(color = "tomato", linewidth = 1.2) +
  geom_point(color = "tomato", size = 2) +
  labs(
    title = "Total Budget by Year",
    x = "Year",
    y = "Budget (in Millions USD)"
  ) +
  theme_minimal()

```

```{r Total Revenue over time}

# Plot 2: Total Revenue
ggplot(yearly_data, aes(x = Year, y = Total_Revenue)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Total Revenue by Year",
    x = "Year",
    y = "Revenue (in Millions USD)"
  ) +
  theme_minimal()

```


Number of Movies Released per Year: Detect spikes or declines in production.

```{r total movie count per year}
## total movie count per year
ggplot(yearly_data, aes(x = Year, y = Movie_count)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Total Movies by Year",
    x = "Year",
    y = "Movie_count"
  ) +
  theme_minimal()
```


# ⭐ Cast Popularity and Impact

## Most Frequent Actors: 
Identify top recurring actors across movies.

```{r top 15 actors by appearances} 
actor_appearance <- clean_movie_data %>%
  select(starts_with("Cast")) %>%
  pivot_longer(cols = everything(), names_to = "Cast_Role", values_to = "Actor") %>%
  filter(!is.na(Actor)) %>%
  group_by(Actor) %>%
  summarise(Appearances = n(), .groups = "drop") %>%
  arrange(desc(Appearances)) %>%
  slice_head(n = 15)

# Plot top 15 actors by appearances
ggplot(actor_appearance, aes(x = fct_reorder(Actor, Appearances), y = Appearances)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Top 15 Most Frequent Actors",
    x = "Actor",
    y = "Number of Movie Appearances"
  ) +
  theme_minimal()

```


## Average Revenue by Actor:

Which actor brings in the highest box office revenue?

```{r top actors by average revenue}
# Merge cast info with revenue
actor_revenue <- clean_movie_data %>%
  select(`Box Office Revenue`, starts_with("Cast")) %>%
  pivot_longer(cols = starts_with("Cast"), names_to = "Cast_Role", values_to = "Actor") %>%
  filter(!is.na(Actor)) %>%
  group_by(Actor) %>%
  summarise(
    Average_Revenue = round(mean(`Box Office Revenue`, na.rm = TRUE) / 1e6, 1),
    Movie_Count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Average_Revenue)) %>%
  filter(Movie_Count >= 3) %>%  # Optional: filter to avoid one-movie anomalies
  slice_head(n = 15)

# Plot top actors by average revenue
ggplot(actor_revenue, aes(x = fct_reorder(Actor, Average_Revenue), y = Average_Revenue)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 15 Actors by Average Revenue (Min 3 Movies)",
    x = "Actor",
    y = "Avg Box Office Revenue (in Millions)"
  ) +
  theme_minimal()
```


# 🎬 Director-Genre Specialization

Which genres do specific directors frequently work in?

```{r visualize top 15 director-genre combos}
# Count how many times each director works in each genre
director_genre_freq <- clean_movie_data %>%
  filter(!is.na(Director_1), !is.na(Genre_1)) %>%
  group_by(Director_1, Genre_1) %>%
  summarise(Movie_Count = n(), .groups = "drop") %>%
  arrange(desc(Movie_Count))

# View top combinations
head(director_genre_freq, 10)

# Optional: visualize top 15 director-genre combos
top_director_genre <- director_genre_freq %>%
  slice_max(order_by = Movie_Count, n = 15)

ggplot(top_director_genre, aes(x = fct_reorder(paste(Director_1, Genre_1, sep = " - "), Movie_Count), y = Movie_Count)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "Top 15 Director–Genre Combinations",
    x = "Director - Genre",
    y = "Movie Count"
  ) +
  theme_minimal()

```


Are certain directors more successful with specific genres?
```{r Compute profit and success metrics by director & genre}
#Compute profit and success metrics by director & genre
director_genre_success <- clean_movie_data %>%
  mutate(
    Profit = `Box Office Revenue` - Budget
  ) %>%
  filter(!is.na(Director_1), !is.na(Genre_1)) %>%
  group_by(Director_1, Genre_1) %>%
  summarise(
    Movies = n(),
    Avg_Revenue = round(mean(`Box Office Revenue`, na.rm = TRUE) / 1e6, 1),
    Avg_Profit = round(mean(Profit, na.rm = TRUE) / 1e6, 1),
    .groups = "drop"
  ) %>%
  filter(Movies >= 3) %>% # optional: only show directors with at least 3 movies in a genre
  arrange(desc(Avg_Profit))

# Top 10 director–genre combos by profit
top_success_director_genre <- director_genre_success %>%
  slice_max(order_by = Avg_Profit, n = 10)

ggplot(top_success_director_genre, aes(x = fct_reorder(paste(Director_1, Genre_1, sep = " - "), Avg_Profit), y = Avg_Profit)) +
  geom_col(fill = "seagreen") +
  coord_flip() +
  labs(
    title = "Top 10 Most Profitable Director–Genre Combos",
    x = "Director - Genre",
    y = "Average Profit (in Millions)"
  ) +
  theme_minimal()
```

# 💸 Profitability Analysis

ROI (Return on Investment) per movie = (Revenue - Budget)/Budget

```{r Plot top ROI movies}
roi_data <- clean_movie_data %>%
  mutate(
    ROI = (`Box Office Revenue` - Budget) / Budget
  ) %>%
  filter(!is.na(ROI), Budget > 0, ROI != Inf) %>%
  arrange(desc(ROI)) %>%
  slice_head(n = 10)  # Top 10 ROI

# Plot top ROI movies
ggplot(roi_data, aes(x = fct_reorder(`Movie Title`, ROI), y = ROI)) +
  geom_col(fill = "darkgreen") +
  geom_text(aes(label = paste0(round(ROI * 100, 0), "%")), hjust = -0.1, size = 3.5) +
  labs(
    title = "Top 10 Movies by ROI",
    x = "Movie Title",
    y = "Return on Investment (ROI)",
    caption = "ROI = (Revenue - Budget) / Budget"
  ) +
  coord_flip() +
  theme_minimal()
```


#  🧑‍🤝‍🧑 Team Effectiveness

## Director–Actor Combos:
Which director-actor pairs consistently generate success?
```{r  Which director-actor pairs consistently generate success?}
director_actor_profit <- clean_movie_data %>%
  select(`Movie Title`, Director_1, starts_with("Cast"), Budget, `Box Office Revenue`) %>%
  mutate(Profit = `Box Office Revenue` - Budget) %>%
  pivot_longer(cols = starts_with("Cast"), names_to = "Cast_Role", values_to = "Actor") %>%
  filter(!is.na(Director_1), !is.na(Actor)) %>%
  group_by(Director_1, Actor) %>%
  summarise(
    Movie_Count = n(),
    Avg_Profit = round(mean(Profit, na.rm = TRUE) / 1e6, 1),
    Success_Rate = mean(Profit > 0),
    .groups = "drop"
  ) %>%
  filter(Movie_Count >= 2)  # Only consider recurring pairs

top_avg_profit_pairs <- director_actor_profit %>%
  arrange(desc(Avg_Profit)) %>%
  slice_head(n = 5)

ggplot(top_avg_profit_pairs, aes(x = fct_reorder(paste(Director_1, Actor, sep = " & "), Avg_Profit), y = Avg_Profit)) +
  geom_col(fill = "purple") +
  geom_text(aes(label = paste0(Avg_Profit, "M")), hjust = -0.1, size = 3.5) +
  labs(
    title = "Top 5 Director–Actor Pairs by Average Profit",
    x = "Director & Actor",
    y = "Avg Profit (in Millions)"
  ) +
  coord_flip() +
  theme_minimal()
```


Top 5 collaborative pairs by average profit or success rate.

```{r}
# Combine all cast columns into one long format
actor_appearance <- clean_movie_data %>%
  select(starts_with("Cast")) %>%
  pivot_longer(cols = everything(), names_to = "Cast_Role", values_to = "Actor") %>%
  filter(!is.na(Actor)) %>%
  group_by(Actor) %>%
  summarise(Appearances = n(), .groups = "drop") %>%
  arrange(desc(Appearances)) %>%
  slice_head(n = 15)

```


# 🎭 Multi-genre Movies
Compare single-genre vs. dual-genre movies in terms of:

Average budget

Average revenue

Success rate

```{r Label movies based on genre columns}
# Label movies based on genre columns
genre_comparison <-movie_data %>%
  mutate(
    Genre_Type = if_else(!is.na(Genre_2), "Multi-genre", "Single-genre"),
    Profit = `Box Office Revenue` - Budget,
    Success = Profit > 0
  ) %>%
  group_by(Genre_Type) %>%
  summarise(
    Avg_Budget = round(mean(Budget, na.rm = TRUE) / 1e6, 1),
    Avg_Revenue = round(mean(`Box Office Revenue`, na.rm = TRUE) / 1e6, 1),
    Success_Rate = round(mean(Success, na.rm = TRUE) * 100, 1),
    Movie_Count = n(),
    .groups = "drop"
  )

print(genre_comparison)

```


# 📈 Genre Revenue Efficiency
Revenue-to-budget ratio per genre.
```{r Revenue-to-budget ratio per genre}
genre_efficiency <- clean_movie_data %>%
  filter(!is.na(Genre_1)) %>%
  group_by(Genre_1) %>%
  summarise(
    Total_Revenue = sum(`Box Office Revenue`, na.rm = TRUE),
    Total_Budget = sum(Budget, na.rm = TRUE),
    Efficiency = round(Total_Revenue / Total_Budget, 2),
    .groups = "drop"
  ) %>%
  filter(Total_Budget > 0) %>%
  arrange(desc(Efficiency))

ggplot(genre_efficiency, aes(x = fct_reorder(Genre_1, Efficiency), y = Efficiency)) +
  geom_col(fill = "darkcyan") +
  geom_text(aes(label = Efficiency), hjust = -0.1, size = 3.5) +
  labs(
    title = "Genre Revenue Efficiency (Revenue-to-Budget Ratio)",
    x = "Genre",
    y = "Revenue / Budget Ratio"
  ) +
  coord_flip() +
  theme_minimal()
```

This tells you which genres are most efficient in generating profit.

# 📆 Release Timing Impact
Analyze by month to see if release timing impacts box office success.

```{r revenue and profit}
release_timing <- clean_movie_data %>%
  mutate(
    Release_Date = mdy_hm(`Release Date`),
    Release_Month = month(Release_Date, label = TRUE, abbr = TRUE),  # e.g., Jan, Feb
    Release_Quarter = quarter(Release_Date, with_year = FALSE),
    Profit = `Box Office Revenue` - Budget,
    Success = Profit > 0
  )

by_month <- release_timing %>%
  group_by(Release_Month) %>%
  summarise(
    Avg_Revenue = round(mean(`Box Office Revenue`, na.rm = TRUE) / 1e6, 1),
    Avg_Profit = round(mean(Profit, na.rm = TRUE) / 1e6, 1),
    Success_Rate = round(mean(Success, na.rm = TRUE) * 100, 1),
    Movie_Count = n(),
    .groups = "drop"
  )

# Revenue
ggplot(by_month, aes(x = Release_Month, y = Avg_Revenue)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Revenue by Release Month", y = "Revenue (in Millions)", x = "Month") +
  theme_minimal()

# Profit
ggplot(by_month, aes(x = Release_Month, y = Avg_Profit)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Average Profit by Release Month", y = "Profit (in Millions)", x = "Month") +
  theme_minimal()
```


# 🧍‍♂️ Ensemble vs. Star-Centric Cast
Compare performance of movies with:

```{r Success Rate by Cast Type}
# Count non-NA cast members
cast_analysis <- movie_data %>%
  mutate(
    Cast_Size = rowSums(!is.na(select(., starts_with("Cast")))),
    Cast_Type = case_when(
      Cast_Size <= 2 ~ "Star-Centric",
      Cast_Size >= 5 ~ "Ensemble",
      TRUE ~ "Mid-Sized"
    ),
    Profit = `Box Office Revenue` - Budget,
    Success = Profit > 0
  )


cast_type_summary <- cast_analysis %>%
  group_by(Cast_Type) %>%
  summarise(
    Avg_Budget = round(mean(Budget, na.rm = TRUE) / 1e6, 1),
    Avg_Revenue = round(mean(`Box Office Revenue`, na.rm = TRUE) / 1e6, 1),
    Avg_Profit = round(mean(Profit, na.rm = TRUE) / 1e6, 1),
    Success_Rate = round(mean(Success, na.rm = TRUE) * 100, 1),
    Movie_Count = n(),
    .groups = "drop"
  )
ggplot(cast_type_summary, aes(x = Cast_Type, y = Success_Rate, fill = Cast_Type)) +
  geom_col() +
  geom_text(aes(label = paste0(Success_Rate, "%")), vjust = -0.5, size = 4) +
  labs(
    title = "Success Rate by Cast Type",
    x = "Cast Type",
    y = "Success Rate (%)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Ensemble" = "darkorange", "Star-Centric" = "skyblue", "Mid-Sized" = "gray"))


```

# Average Revenue by Cast Type

```{r Average Revenue by Cast Type}

ggplot(cast_type_summary, aes(x = Cast_Type, y = Avg_Revenue, fill = Cast_Type)) +
  geom_col() +
  geom_text(aes(label = paste0(Avg_Revenue, "M")), vjust = -0.5, size = 4) +
  labs(
    title = "Average Revenue by Cast Type",
    x = "Cast Type",
    y = "Avg Revenue (in Millions)"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Ensemble" = "darkgreen", "Star-Centric" = "steelblue", "Mid-Sized" = "gray"))

```


# 📌 Conclusion

This analysis provides a comprehensive overview of the key factors influencing movie performance in the dataset. We observed that genres like Horror and Thriller often yield high revenue efficiency, while multi-genre movies and ensemble casts tend to show higher average profits and success rates. Directors and actors with consistent collaborations were identified as strong indicators of commercial success, highlighting the value of effective creative partnerships. Timing of release, particularly in certain quarters, also appears to play a role in box office returns. By integrating genre, cast structure, director-actor dynamics, and release strategy, this report offers actionable insights for understanding what drives profitability and success in the movie industry.
